---
title: "Project 2"
author: "Shashota Saha & Ronit Shah"
date: "2025-03-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr) 
```
#  Introduction

In this project, we aim to identify the best times of day for gardening based on two main data sources:
- Air Quality Data (`daily_44201_2024.csv`) containing ozone levels and AQI measurements.
- Plant Growth Data (`plant_growth_data.csv`) containing sunlight hours, temperature, and plant growth milestones. 

The goal is to find hours where both air quality and plant conditions are optimal.

```{r , echo = TRUE, warning = FALSE , message=FALSE}

```
# Data decription for the Projects
1.daily_44201_2024.csv:

- Description: Contains daily air quality measurements including AQI and ozone levels.

- Important columns: 1st Max Hour, AQI, Arithmetic Mean.

- Covers hourly ozone and air quality patterns across different times of the day.

2. plant_growth_data.csv:

- Description: Experimental dataset on plant growth based on sunlight, temperature, and fertilizer use.

- Important columns: Sunlight_Hours, Temperature, Growth_Milestone.

- Used to find how environmental conditions influence plant health.

```{r , echo=TRUE, warning = FALSE , message=FALSE}

 
```


# Application Functionality:
This analysis includes:

- Cleaning and preprocessing both datasets (dropna used to remove missing data).

- Defining "good gardening hours" based on:

1. Safe air quality (AQI <= 100 and Ozone between 0.030–0.070 ppm).

2. Optimal plant growth conditions (Sunlight 5–8 hours, Temperature 15–30°C, Growth milestone => 1).

- Classifying hours into 4 categories:

1. Best for Plants and Gardeners

2. Good for Plants Only

3. Good for Gardeners Only

4. Bad for Both

- Visualizing these classifications using color-coded histograms and plots. 

# Loading Data 
```{r , echo=TRUE, warning = FALSE , message=FALSE}

# Load Air Quality Data
air_data <- read.csv('daily_44201_2024.csv', stringsAsFactors = FALSE)
air_data <- na.omit(air_data)

# Load Plant Growth Data
plant_data <- read.csv('plant_growth_data.csv', stringsAsFactors = FALSE)
plant_data <- na.omit(plant_data)


head(air_data)
head(plant_data)


```
# Overview of the data 
```{r , echo=TRUE, warning = FALSE , message=FALSE}
summary(air_data)
summary(plant_data)



```
## -Air Data: Gives us average AQI and Ozone by hour.

## -Plant Data: Shows how sunlight, temperature, and growth relate.

# Preprocessing the data 
1. Air Quality Dataset:
We used drop_na() to remove missing AQI, ozone, and hour values for clarity. This ensures accurate classification of air quality based on safe thresholds.

2. Plant Growth Dataset:
Similarly, drop_na() was applied to remove incomplete sunlight, temperature, or growth milestone records.

3. Defining "good gardening hours" based on:

-  Safe air quality (AQI <= 100 and Ozone between 0.030 - 0.070 ppm).

-  Optimal plant growth conditions (Sunlight 5-8 hours, Temperature 15-30°C, Growth milestone >= 1).



# Visual Data Analysis 
```{r , echo=TRUE, warning = FALSE , message=FALSE}

# Scatter plot: AQI vs 1st Max Hour
plot(air_data$X1st.Max.Hour, air_data$AQI,
     main = "Scatter Plot: AQI vs 1st Max Hour",
     xlab = "Hour of Day (0–23)",
     ylab = "Air Quality Index (AQI)",
     pch = 19, col = "blue")


```


```{r , echo=TRUE, warning = FALSE , message=FALSE}
# Scatter plot: Arithmetic Mean vs 1st Max Hour
plot(air_data$X1st.Max.Hour, air_data$Arithmetic.Mean,
     main = "Scatter Plot: Ozone vs 1st Max Hour",
     xlab = "Hour of Day (0–23)",
     ylab = "Ozone (ppm)",
     pch = 19, col = "green")


```

```{r , echo=TRUE, warning = FALSE , message=FALSE}
# Scatter plot: Sunlight Hours vs Growth Milestone
plot(plant_data$Sunlight_Hours, plant_data$Growth_Milestone,
     main = "Scatter Plot: Sunlight Hours vs Growth Milestone",
     xlab = "Sunlight Hours",
     ylab = "Growth Milestone",
     pch = 19, col = "goldenrod")


```

```{r , echo=TRUE, warning = FALSE , message=FALSE}
# Scatter plot: Temperature vs Growth Milestone
plot(plant_data$Temperature, plant_data$Growth_Milestone,
     main = "Scatter Plot: Temperature vs Growth Milestone",
     xlab = "Temperature (°C)",
     ylab = "Growth Milestone",
     pch = 19, col = "red")


```

# Definig Good Gardening Hours



```{r, echo=TRUE, warning=FALSE, message=FALSE}
# Define good gardening time for plants
plant_data$Good_Gardening_Time <- with(plant_data,
  Sunlight_Hours >= 5 & Sunlight_Hours <= 8 &
  Temperature >= 15 & Temperature <= 30 &
  Growth_Milestone >= 1
)

# Filter good plant hours
good_gardening_times <- plant_data[plant_data$Good_Gardening_Time == TRUE, ]
plant_good_hours <- as.numeric(rownames(good_gardening_times))



```

# Define good gardening time for gardeners (based on air quality)
```{r, echo=TRUE, warning=FALSE, message=FALSE}

grouped <- air_data %>%
  group_by(X1st.Max.Hour) %>%
  summarize(AQI = mean(AQI), Arithmetic.Mean = mean(Arithmetic.Mean))

grouped$Good_for_Gardening <- with(grouped,
  AQI <= 100 & Arithmetic.Mean >= 0.030 & Arithmetic.Mean <= 0.070
)

# Filter good gardener hours
good_hours <- grouped[grouped$Good_for_Gardening == TRUE, ]
gardener_good_hours <- good_hours$X1st.Max.Hour

```



# Define good gardening time for gardeners (based on air quality)
```{r, echo=TRUE, warning=FALSE, message=FALSE}

grouped <- air_data %>%
  group_by(X1st.Max.Hour) %>%
  summarize(AQI = mean(AQI), Arithmetic.Mean = mean(Arithmetic.Mean))

grouped$Good_for_Gardening <- with(grouped,
  AQI <= 100 & Arithmetic.Mean >= 0.030 & Arithmetic.Mean <= 0.070
)

# Filter good gardener hours
good_hours <- grouped[grouped$Good_for_Gardening == TRUE, ]
gardener_good_hours <- good_hours$X1st.Max.Hour

```


# Define all 24 hours
```{r, echo=TRUE, warning=FALSE, message=FALSE}
all_hours <- 0:23

# Create label for each hour
hour_labels <- lapply(all_hours, function(hour) {
  if (hour %in% plant_good_hours & hour %in% gardener_good_hours) {
    label <- "Best for Plants and Gardeners"
  } else if (hour %in% plant_good_hours) {
    label <- "Good for Plants Only"
  } else if (hour %in% gardener_good_hours) {
    label <- "Good for Gardeners Only"
  } else {
    label <- "Bad for Both"
  }
  data.frame(Hour = hour, Gardening_Quality = label)
})

# Bind all labels into a table
hour_table <- do.call(rbind, hour_labels)

# Show the final table
hour_table
```


# Conclusion 
This project successfully identifies the best gardening hours based on both plant growth conditions and air quality:

- Best Hours were typically found between mid-morning and early afternoon.

- Bad Hours were more common early morning or late evening when conditions were unfavorable.

- Factors like sunlight, moderate temperature, and clean air together determine optimal gardening time.

This project emphasizes the importance of data cleaning, filtering, and visualization to make practical real-life decisions. Future work can expand this by including humidity, soil quality, or seasonal variations.

# References
-Air Quality Dataset (daily_44201_2024.csv) :U.S. Environmental Protection Agency

-Plant Growth Dataset (plant_growth_data.csv) Experimental Simulation





